Liveness-Based Garbage Collection for Lazy Languages

### 2015-11-26 paper/proof.tex

\input{fh_defs}
\def\drawplusplus#1#2#3{\hbox to 0pt{\hbox to #1{\hfill\vrule height #3 depth
      0pt width #2\hfill\vrule height #3 depth 0pt width #2\hfill
      }}\vbox to #3{\vfill\hrule height #2 depth 0pt width
      #1 \vfill}}
      %Poor man's typography
\def\concat{\mathrel{\drawplusplus {12pt}{0.4pt}{5pt}}}
      %It would be better to specify these in font-relative measures, but it
      %probably doesn't scale anyway.
%\newcommand{\mycomment}[1]{}
\definecolor{Myblue}{rgb}{.2,0,1}
\newcommand{\comment}[1]{{\color{Myblue}{#1}}}
%\newcommand{\cblue}[1]{{\color{blue}{#1}}}
%\newcommand{\blankout}[1]{}

%\newcommand{\deltacall}[3]{\delta_{#1}({#2},{#3})}
\newcommand{\scmin} {\mbox{\sf\em in}}
\newcommand{\scmnull}{\mbox{\sf\em null?}}
\newcommand{\scmpair}{\mbox{\sf\em pair?}}
\newcommand{\scmprim}{\ensuremath{\mathsf{+}}}
\newcommand{\sh}[1]{{\colorbox{gray!20}{\framebox{$#1$}}}}
\def\myvec{\mathaccent"017E } 
\newcommand{\stk}{\mbox{S}}   
\newcommand{\ID}{\mbox{$\mathbf{ id}$}}
\newcommand{\bang}{\mbox{\sc bang}}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{lemma}[theorem]{Lemma}
\begin{document}
% --- Author Metadata here ---
%\conferenceinfo{WOODSTOCK}{'97 El Paso, Texas USA}
%\CopyrightYear{2007} % Allows default copyright year (20XX) to be
%over-ridden - IF NEED BE.
%\crdata{0-12345-67-8/90/01}  % Allows default copyright data
%(0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
% --- End of Author Metadata ---

@@

$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho,\, S,\, H,\, (\CAR~x)
\rightsquigarrow
\rho',\, (\rho, x, (\CAR~x))\!:\!S, \,H,\, s$      &
\sc{car-clo}
\\
\hline

@@

\subsection{Semantics}
\comment{
We now give  a small-step semantics for our  language.
We first specify the domains used by the semantics:
\[
\renewcommand{\arraystretch}{1}
\begin{array}{@{}r@{\ }l@{\ \ }c@{\ \ }l@{\hspace{0.5em}}l}
\heap: & \mathit{Heap} & =&\mathit{Loc} \rightarrow Data & \mbox{-- Heap}\\
d: & \mathit{Data} &=&\mathit{Val} + \mathit{Clo} & \mbox{-- Values \& Closures} \\
v:   & \mathit{Val} &=& \mathbb{N} + \{\NIL\} + \mathit{Data \times Data}& \mbox{-- Values}\\
c:   & \mathit{Clo} &=& \mathit{(App \times Env)}& \mbox{-- Closures}\\
\rho: & \mathit{Env} &=&\mathit{Var} \rightarrow \mathit{Loc} &
\mbox{-- Environment} \\
\end{array}
\]

@@

The  semantics  of   expressions  (and  applications\footnote{In  most
  contexts, we  shall use the  term 'expression' and the  notation $e$
  for both expressions and applications.}) are given by transitions of
the form $\rho, \stk, \heap,  e \rightarrow \rho', \stk', \heap', e'$.
Here \stk\ is a stack of continuation frames.  Each continuation frame
is of the form $(\rho, \ell,  e)$, signifying that the location $\ell$
has  to  be  updated  with  the  value  of  the  currently  evaluating
expression, and $e$ is to be evaluated next in the environment $\rho$.
Note that the  state of the transition system consists  of the current
evaluation  context,  given  by  $\rho$ and  $e$,  and  the  suspended
evaluation contexts (represented as  continuation frames) in the stack
$S$.  The heap $H$ is common to all contexts.

@@

The small-step semantics  is shown in Figure~\ref{fig:lang-semantics}.
As a  sample, consider the  three rules for  $(\CAR~x)$.  If $x$  is a
closure, it is evaluated to WHNF,  say $(d_1, d_2)$.  This is given by
the  rule  {\sc  car-clo}.  If  $d_1$  is  not  in  WHNF, it  is  also
evaluated  ({\sc car-1-clo}). The address ($addr (d_1)$) to be updated with the evaluated
value is  recorded in  a continuation frame.  This is
required for lazy evaluation,
%of an application involving a selector such  as $\CAR$, 
else, in the presence of sharing,   $d_1$  may  be  evaluated  more  than  once~\cite{Jones87}.    Only after this  is the
actual selection done ({\sc car-select}).

}
%==============================================================
\renewcommand{\pp}[2]{\ensuremath{#1\!\!:\!#2}} % prog point
%==============================================================
\begin{figure*}[t]
\renewcommand{\arraystretch}{1}
\begin{eqnarray*}
\mathit{ref\/}(\kappa,\sigma,\Lfonly)
          &=& \{\,\} \mbox{, for $\kappa$ a constant, including
$\NIL$}\\
\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
          &=& \{x.\alpha \mid \acar\alpha \in \sigma\} \cup \{y.\alpha
\mid \acdr\alpha \in \sigma\} \\
\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\} \cup \{x.\acar\alpha \mid \alpha \in
\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\CDR~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\} \cup \{x.\acdr\alpha \mid \alpha \in
\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\PRIM~x~y),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon, y.\epsilon\},  & \mbox{if}~\sigma \ne
\emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\NULLQ~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((f~\myvec{x}),\sigma,\Lfonly)
%          &=& \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
          &=&  \begin{array}{@{}l}  % to discourage \displaystyle
               \bigcup_{i=1}^n x_i.\Lf{f}{i}{\sigma}
               \end{array}
%          &=& \bigcup \{y_i.\Lf{f}{i}{\sigma} \mid i=1,\ldots, n\}
\\[1ex]
\mathcal{L}((\SRETURN~\psi:x),\sigma,\Lfonly) &=& \lbrack \mathit{\epath}
\mapsto
x.\sigma\rbrack, \mbox{ where $\mathit{\epath}$ is a
  new e-path terminating with $\psi$}\\
\mathcal{L}((\SIF~\psi:x~e_1~e_2),\sigma,\Lfonly) &=&
\begin{array}{l l}
                    \mathcal{L}(e_1,\sigma,\Lfonly) \uplus
        \mathcal{L}(e_2,\sigma,\Lfonly) \uplus
        \lbrack\lbrace \mathit{\epath} \mapsto
\{x.\epsilon\}\rbrace\rbrack,  & \mbox{if}~\sigma \ne \emptyset\\
        \emptyset_{\cal{M}}  & \mbox{otherwise}
                 \end{array} \\
&&\mbox{where $\mathit{\epath}$ is a new e-path
  terminating with $\psi$}\\
\mathcal{L}(\LET~x \leftarrow~s~\IN~e),\sigma,\Lfonly) &=&
        \lambda\mathit{\epath}.\; \Lfonly_f(\Lv(x)) \cup \Lv~~
\mbox{where}~\mathcal{M} =
\mathcal{L}(e,\sigma,\Lfonly),~\Lv =
\mathcal{M}(\epath),~\mbox{and}~f~\mbox{is the new function}\\
&& \makebox[0mm]{\hspace*{7cm}
 $(\DEFINE~(f\,x_1\,x_2\, \ldots \, x_n)~(\SIF\,*~x~s[(f\,x_1\,
           x_2\, \ldots\, x_n)/x]))$,} \\
 \makebox[0mm]{\hspace*{8cm} where
     $x_1\, \ldots\, x_n$ are the variables in $s$}
\end{eqnarray*}
\begin{minipage}{0.85\textwidth}
\infrule[live-define]
        {\mathcal{L}(e_f,\sigma,\Lfonly) =
           \bigcup_{i=1}^n z_i.\Lf{f}{i}{\sigma}
              \mbox{ for each $f$ and $\sigma$}
        }
        { \mathit{df_1} \ldots \mathit{df_k} \len \Lfonly
\\ \makebox[0mm]{where
     $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \ e_f)$ is a member of $\mathit{df_1}
\ldots \mathit{df_k}$}}
\end{minipage}
  \caption{Liveness equations and judgement rule}\label{fig:live-judge}
\end{figure*}

@@

We shall now  outline a proof of correctness of  the liveness analysis
presented   in    Section~\ref{sec:liveness-analysis}.    This   proof
technique  is  similar to  the  technique  used in  \cite{asati14lgc}.

The  idea behind  the  proof is  as follows:  We  modify the  standard
semantics to a semantics called  {\em minefield semantics} that models
a liveness based  garbage collection at each  transition.  Before each
transition,  we start  from the  root-set and  insert a  special value
$\bot$ at each  location in the heap that contains  a reference and is
reachable but not live. The proof is based on a context-sensitive form
of    liveness   of    which   our    calculated   liveness    is   an
over-approximation. The state of the transition system is augmented to
carry enough information to calculate  the liveness environment at the
current  evaluation  context  and  each of  the  suspended  evaluation
contexts on  the stack.   To simulate  garbage collection, before each
transition,  we start  from the  root-set and  insert a  special value
$\bot$ at each  location in the heap that contains  a reference and is
reachable but  not live by  our analysis.  Any attempt  to dereference
such locations during the transition will result in entering a special
state denoted \bang.  The main proof is to show that no program enters
the \bang\ state in the minefield semantics.

\newcommand{\cred}[1]{\psframebox[linestyle=none,
    fillcolor=lightgray,fillstyle=solid,framesep=0.5pt]{#1}}

\comment{
\begin{enumerate}
\item Do you need to carry the function body?
\item Minefield doesn't insert bottom inside closures. Separate rule to
be inserted.
\item Examine Proposition 4.1, See that different cases are disjoint
  and together cover all possible cases
\item Suppose the statement that we want to prove is this: Every step
  (step = GC + $\rightsquigarrow$) goes without a bang. How to divide
  a (n+1)-step transition into 1 and n. What is the exact invariant
  required before each step. 
\end{enumerate}

}

>>>>

\input{fh_defs}
\def\drawplusplus#1#2#3{\hbox to 0pt{\hbox to #1{\hfill\vrule height #3 depth
      0pt width #2\hfill\vrule height #3 depth 0pt width #2\hfill
      }}\vbox to #3{\vfill\hrule height #2 depth 0pt width
      #1 \vfill}}
      %Poor man's typography
\def\concat{\mathrel{\drawplusplus {12pt}{0.4pt}{5pt}}}
      %It would be better to specify these in font-relative measures, but it
      %probably doesn't scale anyway.
%\newcommand{\mycomment}[1]{} 
\definecolor{Myblue}{rgb}{.2,0,1}
\definecolor{Myred}{rgb}{1,.2,.2}

\newcommand{\comment}[1]{{\color{Myblue}{#1}}}
\newcommand{\warning}[1]{{\color{Myred}{#1}}}
%\newcommand{\blankout}[1]{}

\newcommand{\cred}[1]{\psframebox[linestyle=none, fillcolor=lightgray,fillstyle=solid,framesep=0.5pt]{#1}}
%\newcommand{\deltacall}[3]{\delta_{#1}({#2},{#3})}
\newcommand{\scmin} {\mbox{\sf\em in}}
\newcommand{\scmnull}{\mbox{\sf\em null?}}
\newcommand{\scmpair}{\mbox{\sf\em pair?}}
\newcommand{\scmprim}{\ensuremath{\mathsf{+}}}
\newcommand{\sh}[1]{{\colorbox{gray!20}{\framebox{$#1$}}}}
\def\myvec{\mathaccent"017E } 
\newcommand{\stk}{\mbox{S}}   
\newcommand{\ID}{\mbox{$\mathbf{ id}$}}
\newcommand{\bang}{\mbox{\sc bang}}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{lemma}[theorem]{Lemma}
\begin{document}
% --- Author Metadata here ---
%\conferenceinfo{WOODSTOCK}{'97 El Paso, Texas USA}
%\CopyrightYear{2007} % Allows default copyright year (20XX) to be
%over-ridden - IF NEED BE.
%\crdata{0-12345-67-8/90/01}  % Allows default copyright data
%(0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
% --- End of Author Metadata ---

@@

$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho,\, S,\, H,\, (\CAR~x)
\rightsquigarrow
\rho',\, (\rho, \rho(x), (\CAR~x))\!:\!S, \,H,\, s$      &
\sc{car-clo}
\\
\hline

@@

\subsection{Semantics}
\comment{
We now give  a small-step semantics for our  language.
We first specify the domains used by the semantics:
\[
\renewcommand{\arraystretch}{1}
\begin{array}{@{}r@{\ }l@{\ \ }c@{\ \ }l@{\hspace{0.5em}}l}
\heap: & \mathit{Heap} & =&\mathit{Loc} \rightarrow Data + \{empty\} & \mbox{-- Heap}\\
d: & \mathit{Data} &=&\mathit{Val} + \mathit{Clo} & \mbox{-- Values \& Closures} \\
v:   & \mathit{Val} &=& \mathbb{N} + \{\NIL\} + \mathit{Data \times Data}& \mbox{-- Values}\\
c:   & \mathit{Clo} &=& \mathit{(App \times Env)}& \mbox{-- Closures}\\
\rho: & \mathit{Env} &=&\mathit{Var} \rightarrow \mathit{Loc} &
\mbox{-- Environment} \\
\end{array}
\]

@@

The  semantics  of   expressions  (and  applications\footnote{In  most
  contexts, we  shall use the  term 'expression' and the  notation $e$
  for both expressions and applications.}) are given by transitions of
the form $\rho, \stk, \heap,  e \rightarrow \rho', \stk', \heap', e'$.
Here \stk\ is a stack of continuation frames.  Each continuation frame
is of the form $(\rho, \ell,  e)$, signifying that the location $\ell$
has  to  be  updated  with  the  value  of  the  currently  evaluating
expression, and $e$ is to be evaluated next in the environment $\rho$.
Note that the  state of the transition system consists  of the current
evaluation  context,  given  by  $\rho$ and  $e$,  and  the  suspended
evaluation contexts (represented as  continuation frames) in the stack
$S$.  The heap $H$ is common to all contexts. The 3-tuple $\rho, \stk,
\heap$ representing 
the state of the transition system will be denoted $\mathcal{S}$.

@@

The small-step semantics  is shown in Figure~\ref{fig:lang-semantics}.
As a  sample, consider the  three rules for  $(\CAR~x)$.  If $x$  is a
closure, it is evaluated to WHNF,  say $(d_1, d_2)$.  This is given by
the  rule  {\sc  car-clo}.  If  $d_1$  is  not  in  WHNF, it  is  also
evaluated  ({\sc car-1-clo}). The address ($addr (d_1)$) to be updated with the evaluated
value is  recorded in  a continuation frame.  This is
required for lazy evaluation,
%of an application involving a selector such  as $\CAR$, 
else, in  the presence of  sharing, $d_1$  may be evaluated  more than
once~\cite{Jones87}.   Only after  this is  the actual  selection done
({\sc car-select}).  $\stackrel{+}{\rightsquigarrow}$
denotes  the  transitive closure  of  $\rightsquigarrow$.

%% We call the sequence of  $\rightsquigarrow$ steps that starts with the
%% evaluation of  an application  $s$ in a  state $\mathcal{S}$  and ends
%% with its  value overwriting  the location of  $s$ as  an \emph{closure
%%   evaluation subtrace (CES)} of $s$ in state $\mathcal{S}$. As an example,
%% an $CES$ starting  with the {\sc car-clo} step may  have the form shown
%% below.   In  this   $H(\rho(x))$  is  assumed  to   be  $\langle  s_1,
%% \rho_1\rangle$ and $d_1$ is $\langle s_2, \rho_2\rangle$. \\

%% 1:$\rho,\, (\_, \ell, \_)\!:\!S,\, H,\, (\CAR~x)
%% \rightsquigarrow$\\
%% \hspace*{.4cm}2:$\rho_1,\, (\rho, \rho(x), (\CAR~x))\!:\!(\_, \ell,
%% \_)\!:\!S, \,H,\, s_1 \stackrel{+}{\rightsquigarrow}$ \\ 
%% \hspace*{.4cm}3:$\rho,\, (\_, \ell, \_)\!:\!S,\, H'[\rho(x)=(d_1, d_2)],\, (\CAR~x)
%%  \rightsquigarrow$\\
%% \hspace*{.4cm}4:$\rho_2,\, (\rho, \rho(x), (\CAR~x))\!:\!(\_, \ell, \_)\!:\!S, \,H',\,
%% s_2 \stackrel{+}{\rightsquigarrow}$\\
%% \hspace*{.4cm}5:$\rho,\, (\_, \ell, \_)\!:\!S,\, H''[\rho(x)=(v_1, d_2)],\, (\CAR~x)
%% \rightsquigarrow$\\ 
%% \hspace*{.4cm}6:$\rho,\, S,\, H''[\ell=(v_1, d_2)],\, (\CAR~x)$ \\

%% Notice that the $CES$ that begins  with a {\sc car-clo} is composed of
%% four  more $CES$s  that  start at  lines 2  (evaluation  of $s_1$),  3
%% (evaluation of $(\CAR~x)$ in a different state, 4 (evaluation of $s_2$
%% and 5 (evaluation of $(\CAR~x)$ in a yet different state).

}
%==============================================================
\renewcommand{\pp}[2]{\ensuremath{#1\!\!:\!#2}} % prog point
%==============================================================
\begin{figure*}[t]
\comment{
\renewcommand{\arraystretch}{1}
\begin{eqnarray*}
\mathit{ref\/}(\kappa,\sigma,\Lfonly)
          &=& \{\,\} \mbox{, for $\kappa$ a constant, including
$\NIL$}\\
\mathit{ref\/}(\pi:(\CONS~x~y),\sigma,\Lfonly)
          &=& \{x_{\pi}.\alpha \mid \acar\alpha \in \sigma\} \cup
  \{y_{\pi}.\alpha 
\mid \acdr\alpha \in \sigma\} \\
\mathit{ref\/}(\pi:(\CAR~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x_{\pi}.\epsilon\} \cup \{x_{\pi}.\acar\alpha \mid \alpha \in
\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}(\pi:(\CDR~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x_{\pi}.\epsilon\} \cup \{x_{\pi}.\acdr\alpha \mid \alpha \in
\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}(\pi:(\PRIM~x~y),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x_{\pi}.\epsilon, y_{\pi}.\epsilon\},  & \mbox{if}~\sigma \ne
\emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}(\pi:(\NULLQ~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x_{\pi}.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}(\pi:(f~\myvec{x}),\sigma,\Lfonly)
%          &=& \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
          &=&  \begin{array}{@{}l}  % to discourage \displaystyle
               \bigcup_{i=1}^n x_{i_{\pi}}.\Lf{f}{i}{\sigma}
               \end{array}
%          &=& \bigcup \{y_i.\Lf{f}{i}{\sigma} \mid i=1,\ldots, n\}
\\[1ex]
\mathcal{L}((\SRETURN~\psi:x),\sigma,\Lfonly) &=& \lbrack \mathit{\epath}
\mapsto
x.\sigma\rbrack, \mbox{ where $\mathit{\epath}$ is a
  new e-path terminating with $\psi$}\\
\mathcal{L}((\SIF~\psi:x~e_1~e_2),\sigma,\Lfonly) &=&
\begin{array}{l l}
                    \mathcal{L}(e_1,\sigma,\Lfonly) \uplus
        \mathcal{L}(e_2,\sigma,\Lfonly) \uplus
        \lbrack\lbrace \mathit{\epath} \mapsto
\{x.\epsilon\}\rbrace\rbrack,  & \mbox{if}~\sigma \ne \emptyset\\
        \emptyset_{\cal{M}}  & \mbox{otherwise}
                 \end{array} \\
&&\mbox{where $\mathit{\epath}$ is a new e-path
  terminating with $\psi$}\\
%% \mathcal{L}((\LET~x \leftarrow~s~\IN~e),\sigma,\Lfonly) &=&
%%         \lambda\mathit{\epath}.\; \Lfonly_f(\Lv(x)) \cup \Lv~~
%% \mbox{where}~\mathcal{M} =
%% \mathcal{L}(e,\sigma,\Lfonly),~\Lv =
%% \mathcal{M}(\epath),~\mbox{and}~f~\mbox{is the new function}\\
%% && \makebox[0mm]{\hspace*{7cm}
%%  $(\DEFINE~(f\,x_1\,x_2\, \ldots \, x_n)~(\SIF\,*~x~s[(f\,x_1\,
%%            x_2\, \ldots\, x_n)/x]))$,} \\
%%  \makebox[0mm]{\hspace*{8cm} where
%%      $x_1\, \ldots\, x_n$ are the variables in $s$}
\mathcal{L}((\LET~x \leftarrow~s~\IN~e),\sigma,\Lfonly) &=&
        \lambda\mathit{\epath}.\; 
\mathit{ref\/}(s,\sigma,\Lfonly) \cup \Lv\\
&&\mbox{where}~ \mathcal{M} = \mathcal{L}(e,\sigma,\Lfonly), ~\Lv =
\mathcal{M}(\epath),~\mbox{and}~\sigma = \bigcup_{{\pi}\in \epath} \Lv(x_{\pi})
\end{eqnarray*}
\begin{minipage}{0.85\textwidth}
\infrule[live-define]
        {\mathcal{L}(e_f,\sigma,\Lfonly) =
           \bigcup_{i=1}^n z_i.\Lf{f}{i}{\sigma}
              \mbox{ for each $f$ and $\sigma$}
        }
        { \mathit{df_1} \ldots \mathit{df_k} \len \Lfonly
\\ \makebox[0mm]{where
     $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \ e_f)$ is a member of $\mathit{df_1}
\ldots \mathit{df_k}$}}
\end{minipage}
  \caption{Liveness equations and judgement
    rule}\label{fig:live-judge}
}%comment
\end{figure*}

@@

\warning{The (hopefully fool proof) full proof is now being developed
  in the Appendix.}

%% We shall now  outline a proof of correctness of  the liveness analysis
%% presented   in    Section~\ref{sec:liveness-analysis}.    This   proof
%% technique  is  similar to  the  technique  used in  \cite{asati14lgc}.

%% The  idea behind  the  proof is  as follows:  We  modify the  standard
%% semantics to a semantics called  {\em minefield semantics} that models
%% a liveness based  garbage collection at each  transition.  Before each
%% transition,  we start  from the  root-set and  insert a  special value
%% $\bot$ at each  location in the heap that contains  a reference and is
%% reachable but not live. The proof is based on a context-sensitive form
%% of    liveness   of    which   our    calculated   liveness    is   an
%% over-approximation. The state of the transition system is augmented to
%% carry enough information to calculate  the liveness environment at the
%% current  evaluation  context  and  each of  the  suspended  evaluation
%% contexts on  the stack.   To simulate  garbage collection, before each
%% transition,  we start  from the  root-set and  insert a  special value
%% $\bot$ at each  location in the heap that contains  a reference and is
%% reachable but  not live by  our analysis.  Any attempt  to dereference
%% such locations during the transition will result in entering a special
%% state denoted \bang.  The main proof is to show that no program enters
%% the \bang\ state in the minefield semantics.

%% \newcommand{\cred}[1]{\psframebox[linestyle=none,
%%     fillcolor=lightgray,fillstyle=solid,framesep=0.5pt]{#1}}

%% \comment{
%% \begin{enumerate}
%% \item Do you need to carry the function body?
%% \item Minefield doesn't insert bottom inside closures. Separate rule to
%% be inserted.
%% \item Examine Proposition 4.1, See that different cases are disjoint
%%   and together cover all possible cases
%% \item Suppose the statement that we want to prove is this: Every step
%%   (step = GC + $\rightsquigarrow$) goes without a bang. How to divide
%%   a (n+1)-step transition into 1 and n. What is the exact invariant
%%   required before each step. 
%% \end{enumerate}

%% }

%% \section{Minefield semantics for the language}
%% \begin{figure*}[ht!]
%% \begin{center}\footnotesize
%% \begin{tabular}{|c|c|c|}
%% \hline
%% Premise & Transition & Rule name \\ 
%% \hline
%% \hline
%%           &\makecell{ $\rho, (\rho', x, e, \cred{\sigma', e_f'})\!:\!S,
%%   H, \kappa, \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', S, H[\rho'(x) :=
%%     \kappa], e, \cred{\sigma', e_f'}$ }   &  \sc{const}
%% \\
%% \hline
%%           & \makecell{$\rho, (\rho', z, e, \cred{\sigma',
%%     e_f'})\!:\!S, H, (\CONS~x~y), \cred{\sigma, e_f}$ \\ $\rightsquigarrow
%%   \rho', S, H[\rho'(z) := (\rho(x),\rho(y))], e, \cred{\sigma', e_f'}$}     &  \sc{cons} \\
%% \hline
%% $H(\rho(x)) \mbox{ is } (l_1, l_2)$ & \makecell{$\rho, (\rho', z, e,
%%   \cred{\sigma', e_f'} )\!:\!S, H, (\CAR~x), \cred{\sigma, e_f}$ \\ $
%%   \rightsquigarrow \rho', S, H[\rho'(z) := H(l_1)], e, \cred{\sigma', e_f'}$}      &
%% \sc{car-whnf} \\
%% \hline
%% $H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S,
%%   H, (\CAR~x), \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', (\rho, x,
%%   (\CAR~x), \cred{\sigma, e_f})\!:\!S, H, s, \cred{(\clazy \cup \acar)\sigma , \_ }$}      &
%% \sc{car-clo}
%% \\
%% \hline
%% $H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
%%  & \makecell{$\rho, (\rho', z, e, \cred{\sigma', e_f'})\!:\!S, H,
%%   (+~x~y), \cred{\sigma, e_f}$ \\  $\rightsquigarrow \rho', S, H[\rho'(z)
%%     := H(\rho'(x)) + H(\rho'(y))], e, \cred{\sigma', e_f'}$}      &
%% \sc{prim-whnf} \\
%% \hline
%% $H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{$\rho, S,
%%   H, (+~x~y), \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', (\rho, x,
%%   (+~x~y), \cred{\sigma, e_f})\!:\!S, H, s, \cred{\clazy\sigma, \_}$}      &
%% \sc{prim-1-clo} \\
%% \hline
%% $H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & \makecell{$\rho,
%%   S, H, (+~x~y), \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', (\rho, y,
%%   (+~x~y), \cred{\sigma, e_f})\!:\!S, H, s, \cred{\clazy\sigma, \_ }$}      &
%% \sc{prim-2-clo} \\
%% \hline
%% {$\mathit{g}~\mbox{defined as}$
%% $~(\DEFINE~(g~\myvec{y})~e_{\mathit{g}})$}  & \makecell{$\rho, S, H,
%%   (g~\myvec{x}), \cred{\sigma, e_f}$ \\ $\rightsquigarrow [\myvec{y} \mapsto
%%     \rho(\myvec{x})], S, H, e_{\mathit{g}}, \cred{\sigma, e_g}$}      &
%% \sc{funcall} \\
%% \hline
%% $\ell$ is a new location& \makecell{$\rho, S, H, (\LET~x\leftarrow
%%   s~\IN~e), \cred{\sigma, e_f}$ \\ $ \rightsquigarrow \rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s, \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
%%   \ell]\rangle], e, \cred{\sigma, e_f}$} &
%% \sc{let} \\
%% \hline
%% $H(\rho(x)) \ne 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2),
%%   \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho, S, H,  e_1, \cred{\sigma, e_f}$} & \sc{if-true} \\
%% \hline
%% $H(\rho(x)) = 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2),
%%   \cred{\sigma, e_f}$ \\   $\rightsquigarrow
%% \rho, S, H,  e_2, \cred{\sigma, e_f}$} & \sc{if-false} \\
%% \hline
%% $H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,
%%   (\pi:\SIF~\psi:x~e_1~e_2), \cred{\sigma, e_f}$ \\ $\rightsquigarrow
%% \rho', (\rho, x, (\SIF~x~e_1~e_2),  \cred{\sigma, e_f})\!:\!S, H, s,
%% \cred{\clazy\sigma, e_f}$}
%% &
%% \sc{if-clo} \\
%% \hline
%% {$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& \makecell{$\rho,
%%   (\rho', z, e, \cred{\sigma', e_f'})\!:\!S, H, (\SRETURN~x), \cred{\sigma,
%%   e_f}$ \\ $\rightsquigarrow \rho', S, H[\rho'(z) := v], e,
%%   \cred{\sigma', e_f'}$} &
%% \sc{return-whnf}\\
%% \hline
%% $H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,
%%   (\psi:\SRETURN~x), \cred{\sigma, e_f}$ \\ $
%%   \rightsquigarrow$
%% $\rho',~ (\rho, x, (\SRETURN~x), \cred{\sigma, e_f})\!:\!S, H,  s,
%%   \cred{\sigma, e_f}$} &
%% \sc{return-clo} \\
%% \hline
%% \end{tabular}
%% \caption{Minefield semantics.\label{fig:minefield-semantics}}
%% \end{center}
%% \vskip -5mm
%% \end{figure*}

%% To set up the minefield semantics, we follow these steps:
%% \begin{enumerate}
%% \item We enrich  the abstract machine state $\rho,$ $S,$  $ H,$ $e$ to
%%   $\rho,$ $S,$ $H,$ $e,$ $\sigma,$ $e_f$, where $\sigma$ is the demand
%%   on the expression $e$, and $e_f$ is the body of the function. As the
%%   section on liveness  analysis shows, the entire  function body $e_f$
%%   and $\sigma$  are needed  to calculate  the liveness  environment at
%%   different program  points. The  demand sigma  is ``dynamic''  in the
%%   sense  that it  arises from  the actual  function calls  made during
%%   execution.   The  0-CFA  demand  is an  over-approximation  of  this
%%   demand.  The information  in a suspended evaluation  context is also
%%   similarly augmented with  the demand on it and the  function body of
%%   which the suspended evaluation is a  part.  Thus a stacked entry now
%%   takes the  form $(\rho, x, e,  \sigma, e_f)$. A modification  of the
%%   small step semantics which carries the extra information is shown in
%%   Figure~\ref{fig:minefield-semantics}.
%% \item Given  the current  evaluation context $\rho,  S, H,  e, \sigma,
%%   e_f$ and $\mathcal{LF}$, we can construct a liveness environment as follows:
%%   \begin{enumerate}
%%   \item If $e$ is an expression, then:  
%%   $$ \Lv = \bigcup_{\psi \in \mathcal{P}(e)} \mathcal{M}(\psi),
%%     \text{where }
%%   \mathcal{M} = \mathcal{L}(e_f, \sigma, \Lfonly)$$
%% \item If $e$ is an application, then:  
%%   $$ \Lv = ref(e, \sigma, \Lfonly)$$
%%   \end{enumerate}
%% We can define in the same way the liveness environment for each of the
%% suspended  evaluation contexts in  $S$  giving  a  stack of  liveness
%% environments that we shall denote $\mathsf{SL}$.
%% \item $GC$ models a liveness based garbage collection:
%% $GC(\rho,$ $ S,$ $ H,$ $ \mathcal{L},$ $ \mathsf{SL}) = (\rho', S', H')$ 
%%   \begin{enumerate}
%%   \item For each $x \in domain(\rho)$,  $\rho'(x)=\bot$ iff $x.\epsilon
%%     \notin \Lv$.
%% \item For each stack entry $(\rho,\_,\_,\_,\_)$ in $S$ with $\Lv'$ as
%%   the corresponding liveness environment in $\mathsf{SL}$,  and for each $x
%%   \in domain(\rho)$,  $\rho'(x)=\bot$
%%   iff $x.\epsilon \notin \Lv'$.
%% \item For each location $\ell$, $H'(\ell) = \bot$, iff  there is no
%%   $x$ in either the current environment or one of the stacked
%%   environments $\rho$ such that for some $\alpha$, $H[x.\alpha] =
%%   \ell$  and $x.\alpha \in \Lv$, .
%%   \end{enumerate}
%% \end{enumerate}
